\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2022-04-22 submission/manuscript.tex
update manuscript w/revisions

\usepackage{lineno}

\usepackage{pdfpages}
\floatplacement{figure}{H} % Keep the figure up top of the page
\DeclareUnicodeCharacter{0301}{*************************************}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

@@

\pagenumbering{arabic}
\linenumbers
\doublespacing

@@

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether there is an association between members of the gut
microbiota and disease severity. The gut microbiota is known to interact
with \emph{C. difficile} during infection. Perturbations to the gut
microbiota are necessary for \emph{C. difficile} to colonize the gut.
The gut microbiota can inhibit \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
Here we sought to demonstrate that members of the gut bacterial
communities can also contribute to disease severity. We derived diverse
gut communities by colonizing germ-free mice with different human fecal
communities. The mice were then infected with a single \emph{C.
difficile} ribotype 027 clinical isolate which resulted in moribundity
and histopathologic differences. The variation in severity was
associated with the human fecal community that the mice received.
Generally, bacterial populations with pathogenic potential, such as
\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
associated with more severe outcomes. Bacterial groups associated with
fiber degradation, bile acid metabolism and lantibiotic production, such
as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
less severe outcomes. These data indicate that, in addition to the host
and \emph{C. difficile}, populations of gut bacteria can influence CDI
disease severity.

@@

Missing from CDI severity prediction models are the effects of the
indigenous gut bacteria. \emph{C. difficile} interacts with the gut
community in many ways. The indigenous bacteria of a healthy intestinal
community provide a protective barrier preventing \emph{C. difficile}
from infecting the gut. A range of mechanisms can disrupt this barrier,
including antibiotics, medications, or dietary changes, and lead to
increased susceptibility to CDI (4--6). Once \emph{C. difficile}
overcomes the protective barrier and colonizes the intestine, the
indigenous bacteria can either promote or inhibit \emph{C. difficile}
through producing molecules or modifying the environment (7, 8). Bile
acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
growth and affect toxin production (9, 10). Bacteria in the gut also can
compete more directly with \emph{C. difficile} through antibiotic
production or nutrient consumption (11--13). While the relationship
between the gut bacteria and \emph{C. difficile} has been established,
the effect the gut bacteria can have on CDI disease severity is unclear.

Recent studies have demonstrated that when mice with diverse microbial
communities were challenged with a high-toxigenic strain resulted in
varied disease severity (14) and when challenged with a low-toxigenic
strain members of the gut microbial community associated with variation
in colonization (15). Here, we sought to further elucidate the
relationship between members of the gut bacterial community and CDI
disease severity when challenged with a high-toxigenic strain, \emph{C.
difficile} ribotype 027 (RT027). We hypothesized that since specific
groups of gut bacteria affect the metabolism of \emph{C. difficile} and
its infection dynamics, we can also identify groups of bacteria that
affect the disease severity of the infection. To test this hypothesis,
we colonized germ-free C57BL/6 mice with human fecal samples to create
varied gut communities. We then challenged the mice with \emph{C.
difficile} RT027 and followed the mice for the development of severe
outcomes of moribundity and histopathologic cecal tissue damage. Since
the murine host and \emph{C. difficile} isolate were the same and only
the gut community varied, the variation in disease severity we observed
was attributable to the gut microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics.} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (16). We inoculated germ-free C57BL/6 mice with homogenized feces
from each of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (17). The gut communities were
allowed to equilibrate for two weeks post-inoculation (18). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (19, 20). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, such as recent
hospitalization or antibiotic use (17), we tested whether \emph{C.
difficile} was able to infect these mice without an antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with a
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate spores. The mice were followed for 10
days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N1, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N3, N4, and N5, were not colonized at detectable levels on
the day after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (\emph{P} = 0.003). However, when looking at
each group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N2 and N5 through N9 had comparable toxin activity as the
moribund mice. Additionally, not all moribund mice had toxin activity
detected in their stool. Next, we examined the cecal tissue for
histopathologic damage. Moribund mice had high levels of epithelial
damage, tissue edema, and inflammation (Figure S1) similar to previously
reported histopathologic findings for \emph{C. difficile} RT027 (21). As
observed with toxin activity, the moribund mice had higher
histopathologic scores than the non-moribund mice (\emph{P} \textless{}
0.001). However, unlike the toxin activity, all moribund mice had
consistently high histopathologic summary scores (Figure 3B). The
non-moribund mice, Donor groups N1 through N9, had a range in tissue
damage from none detected to similar levels as the moribund mice, which
grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the donor
groups but were largely consistent within each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on their moribundity and histopathologic score. We
dichotomized the histopathologic scores into high and low groups by
splitting on the median score of 5. This analysis revealed 20 genera
that were significantly different by the disease severity (Figure 4A).
Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
\emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher relative abundances
in the mice that became moribund. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
more abundant in the non-moribund mice that would develop only low
intestinal injury. To understand the role of toxin activity in disease
severity, we applied LEfSe to identify the genera most likely to explain
the differences between the presence and absence of detected toxin
activity (Figure 4B). Many genera that associated with the presence of
toxin were also associated with moribundity, such as populations of
\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
\emph{Murimonas} that were associated with no detected toxin that also
exhibited greater relative abundance in communities from non-moribund
mice with a low histopathologic score. Lastly, we tested for
correlations between the endpoint relative abundances of bacterial
operational taxonomic units (OTUs) and the histopathologic summary score
(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
positively correlated with histopathologic score, as its day 0 relative
abundance did with disease severity (Figure 4A). Populations of
\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
with the histopathologic score and were increased in the group of mice
with detectable toxin. This analysis identified bacterial genera that
were associated with the variation in moribundity, histopathologic
score, and toxin.

We next determined whether, collectively, bacterial community membership
and relative abundance could be predictive of the CDI disease outcome.
We trained random forest models with bacterial community relative
abundance data from the day of colonization at each taxonomic rank to
predict toxin, moribundity, and day 10 post-challenge histopathologic
summary score. For predicting if detectable toxin would be produced,
microbial populations aggregated by phylum rank classification performed
similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure
S2). \emph{C. difficile} was more likely to produce detectable toxin
when the community infected had less abundant populations of
\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
ability of the community to predict moribundity. Bacteria grouped by
class rank classification was sufficient to predict which mice would
succumb to the infection before the end of the experiment (AUROC = 0.91,
Figure S2). The features with the greatest effect showed that
communities with greater populations of bacteria belonging to
\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
\emph{Erysipelotrichia} were more likely to result in moribundity
(Figure 5B). Only one other class of bacteria was decreased in moribund
mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
abundances of genera were able to predict a high or low histopathologic
score (histopathologic scores were dichotomized as in previous analysis,
AUROC = 0.99, Figure S2). No genera had a significantly greater effect
on the model performance than any others, indicating the model was
reliant on many genera for the correct prediction. The model used some
of the genera identified in the LEfSe analysis, such as
\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
Communities with greater abundances of \emph{Hungatella},
\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
\emph{Neisseria} were more likely to have high histopathologic scores.
These models have shown that the relative abundance of bacterial
populations and their relationship to each other could be used to
predict the variation in moribundity, histopathologic score, and
detectable toxin of CDI.

@@

Challenging mice colonized with different human fecal communities with
\emph{C. difficile} RT027 demonstrated that variation in members of the
gut microbiome affects \emph{C. difficile} infection disease severity.
Our analysis revealed an association between the relative abundance of
bacterial community members and disease severity. Previous studies
investigating the severity of CDI disease involving the microbiome have
had limited ability to interrogate this relationship between the
microbiome and disease severity. Studies that have used clinical data
have limited ability to control variation in the host, microbiome or
\emph{C. difficile} ribotype (22). Murine experiments typically use a
single mouse colony and different \emph{C. difficile} ribotypes to
create severity differences (23). Recently, our group has begun
uncovering the effect microbiome variation has on \emph{C. difficile}
infection. We showed the variation in the bacterial communities between
mice from different mouse colonies resulted in different clearance rates
of \emph{C. difficile} (15). We also showed varied ability of mice to
spontaneously eliminate \emph{C. difficile} infection when they were
treated with different antibiotics prior to \emph{C. difficile}
challenge (24). Overall, the results presented here have demonstrated
that the gut bacterial community contributed to the severity of \emph{C.
difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 25, 26). Those levels are driven by the
activities in the intestine (27). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (19, 28--31).
However, other studies have shown that disease is not necessarily linked
with toxin production (32) or the strain (33). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (34--37). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
23, 38, 39). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy murine gut
community and is only able to become established after a perturbation
(20). Once colonized, the different communities lead to different
metabolic responses and dynamics of the \emph{C. difficile} population
(8, 24, 40). Gut bacteria metabolize primary bile acids into secondary
bile acids (41, 42). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (9, 10, 43,
44). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (45, 46). The nutrient environment
affects toxin production (47). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

A myriad of studies have explored the relationship between the
microbiome and CDI disease. Studies examining difference in disease
often use different \emph{C. difficile} strains or ribotypes in mice
with similar microbiota as a proxy for variation in disease, such as
strain 630 for non-severe and RT027 for severe (19, 28, 29, 48). Studies
have also demonstrated variation in infection through tapering
antibiotic dosage (20, 24, 49) or by reducing the amount of \emph{C.
difficile} cells or spores used for the challenge (19, 49). These
studies often either lack variation in the initial microbiome or have
variation in the \emph{C. difficile} infection itself, confounding any
association between variation in severity and the microbiome. Recent
studies have shown variation in the initial microbiome, via different
murine colonies or colonizing germ-free mice with human feces, that were
challenged with \emph{C. difficile} resulted in varied outcomes of the
infection (14, 15).

Our data have demonstrated gut bacterial relative abundances associate
with variation in toxin production, histopathologic scoring of the cecal
tissue and mortality. This analysis revealed populations of
\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
more abundant in the microbiome of non-moribund mice which had low
histopathologic scores and no detected toxin. The protective role of
these genera are supported by previous studies. \emph{Coprobacillus},
\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
involved in dietary fiber fermentation and associated with healthy
communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
which produce short chain fatty acids, have been associated with healthy
communities (54--56). Furthermore, \emph{Coprobacillus}, which was
abundant in mice with low histopathologic scores but rare in all other
mice, has been shown to contain a putative type I lantibiotic gene
cluster and inhibit \emph{C. difficile} colonization (57--59).
\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
abundant in mice which had a low histopathologic scores but have
contradictory supporting evidence in the current literature. In our
data, \emph{Akkermansia} was most abundant in the non-moribund mice with
low histopathologic scores but there were some moribund mice which had
increased populations of \emph{Akkermansia}. This could be attributed to
either a more protective mucus layer was present inhibiting colonization
(59, 60) or mucus consumption by \emph{Akkermansia} could have been
crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
and associated with low histopathologic scores. It has been associated
with healthy populations and has been used to mono-colonize germ-free
mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
However, \emph{Enterocloster} has also been involved in infections, such
as bacteremia (66, 67). These data have exemplified populations of
bacteria that have the potential to be either protective or harmful.
Thus, the disease outcome is not likely based on the abundance of
individual populations of bacteria, rather it is the result of the
interactions of the community.

>>>>

\usepackage{lineno}
\pagenumbering{arabic}
\linenumbers

\usepackage{pdfpages}
\floatplacement{figure}{H} % Keep the figure up top of the page
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

@@

@@

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether specific members of the gut microbiota associate with
variation in disease severity. The gut microbiota is known to interact
with \emph{C. difficile} during infection. Perturbations to the gut
microbiota are necessary for \emph{C. difficile} to colonize the gut.
The gut microbiota can inhibit \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
Here we sought to demonstrate that members of the gut bacterial
communities can also contribute to disease severity. We derived diverse
gut communities by colonizing germ-free mice with different human fecal
communities. The mice were then infected with a single \emph{C.
difficile} ribotype 027 clinical isolate which resulted in moribundity
and histopathologic differences. The variation in severity was
associated with the human fecal community that the mice received.
Generally, bacterial populations with pathogenic potential, such as
\emph{Enterococcus}, \emph{Helicobacter}, and \emph{Klebsiella}, were
associated with more severe outcomes. Bacterial groups associated with
fiber degradation and bile acid metabolism, such as \emph{Anaerotignum},
\emph{Blautia}, \emph{Lactonifactor}, and \emph{Monoglobus}, were
associated with less severe outcomes. These data indicate that, in
addition to the host and \emph{C. difficile} subtype, populations of gut
bacteria can influence CDI disease severity.

@@

Missing from CDI severity prediction models are the effects of the
indigenous gut bacteria. \emph{C. difficile} interacts with the gut
community in many ways. The indigenous bacteria of a healthy intestinal
community prevent \emph{C. difficile} from infecting the gut (4). A
range of mechanisms can disrupt this inhibition, including antibiotics,
medications, or dietary changes, and lead to increased susceptibility to
CDI (5--7). Once \emph{C. difficile} overcomes the inhibition and
colonizes the intestine, the indigenous bacteria can either promote or
inhibit \emph{C. difficile} through producing molecules or modifying the
environment (8, 9). Bile acids metabolized by the gut bacteria can
inhibit \emph{C. difficile} growth and affect toxin production (4, 10,
11). Bacteria in the gut also can compete more directly with \emph{C.
difficile} through antibiotic production or nutrient consumption
(12--14). While the relationship between the gut bacteria and \emph{C.
difficile} has been established, the effect the gut bacteria can have on
CDI disease severity is unclear.

Recent studies have demonstrated that when mice with diverse microbial
communities were challenged with a high-toxigenic strain resulted in
varied disease severity (15) and when challenged with a low-toxigenic
strain members of the gut microbial community associated with variation
in colonization (16). Here, we sought to further elucidate the
relationship between members of the gut bacterial community and CDI
disease severity when challenged with a high-toxigenic strain, \emph{C.
difficile} ribotype 027 (RT027). We hypothesized that since specific
groups of gut bacteria affect the metabolism of \emph{C. difficile} and
its clearance rate, specific groups of bacteria associate with variation
in CDI disease severity. To test this hypothesis, we colonized germ-free
C57BL/6 mice with human fecal samples to create varied gut communities.
We then challenged the mice with \emph{C. difficile} RT027 and followed
the mice for the development of severe outcomes of moribundity and
histopathologic cecal tissue damage. Since the murine host and \emph{C.
difficile} isolate were the same and only the gut community varied, the
variation in disease severity we observed was attributable to the gut
microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics.} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (17). We inoculated germ-free C57BL/6 mice with homogenized feces
from each of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (18). The gut communities were
allowed to colonize for two weeks post-inoculation (19). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (20, 21). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, such as recent
hospitalization or antibiotic use (18), we tested whether \emph{C.
difficile} was able to infect these mice without an antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with a
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate spores. The mice were followed for 10
days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N1, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N3, N4, and N5, were not colonized at detectable levels on
the day after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (Figure S1). However, when looking at each
group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N2 and N5 through N9 had comparable toxin activity as the
moribund mice at 2 days post-challenge. Additionally, not all moribund
mice had toxin activity detected in their stool. Next, we examined the
cecal tissue for histopathologic damage. Moribund mice had high levels
of epithelial damage, tissue edema, and inflammation (Figure S2) similar
to previously reported histopathologic findings for \emph{C. difficile}
RT027 (22). As observed with toxin activity, the moribund mice had
higher histopathologic scores than the non-moribund mice (\emph{P}
\textless{} 0.001). However, unlike the toxin activity, all moribund
mice had consistently high histopathologic summary scores (Figure 3B).
The non-moribund mice, Donor groups N1 through N9, had a range in tissue
damage from none detected to similar levels as the moribund mice, which
grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the donor
groups but were largely consistent within each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on moribundity or 10 days post infection (dpi)
histopathologic score for non-moribund. This analysis revealed bacterial
operational taxonomic units (OTUs) that were significantly different at
the time of challenge by the disease severity (Figure 4A). OTUs
associated with \emph{Akkermansia}, \emph{Bacteroides},
\emph{Clostridium sensu stricto}, and \emph{Turicibacter} were detected
at higher relative abundances in the mice that became moribund. OTUs
associated with \emph{Anaerotignum}, \emph{Enterocloster}, and
\emph{Murimonas} were more abundant in the non-moribund mice that would
develop only low intestinal injury. To understand the role of toxin
activity in disease severity, we applied LEfSe to identify the OTUs at
the time of challenge that most likely explain the differences between
communities that had toxin activity detected at anytime point to those
that did not (Figure 4B). An OTU associated with \emph{Bacteroides}, OTU
7, associated with the presence of toxin also associated with
moribundity. Likewise, OTUs associated with \emph{Enterocloster} and
\emph{Murimonas} that were associated with no detected toxin also
exhibited greater relative abundance in communities from non-moribund
mice with a low histopathologic score. Lastly, we tested for
correlations between the endpoint (10 dpi) relative abundances of OTUs
and the histopathologic summary score (Figure 4C). The endpoint relative
abundance of \emph{Bacteroides}, OTU 17, was positively correlated with
histopathologic score, as its day 0 relative abundance did with disease
severity (Figure 4A). A population of \emph{Bacteroides}, OTU 17, was
positively correlated with the histopathologic score and were increased
in the group of mice with detectable toxin. We also tested for
correlations between the endpoint relative abundances of OTUs and toxin
activity but none were significant. This analysis identified bacterial
populations that were associated with the variation in moribundity,
histopathologic score, and toxin.

We next determined whether, collectively, bacterial community membership
and relative abundance could be predictive of the CDI disease outcome.
We trained logistic regression models with bacterial community relative
abundance data from the day of colonization at each taxonomic rank to
predict toxin, moribundity, and histopathologic summary score. For
predicting if detectable toxin would be produced, microbial populations
aggregated by genus rank classification performed similarly as models
using lower taxonomic ranks (mean AUROC = 0.787, Figure S3). \emph{C.
difficile} increased odds of producing detectable toxin when the
community infected had less abundant populations of \emph{Monoglobus},
\emph{Akkermansia}, \emph{Extibacter}, \emph{Intestinimonas} and
\emph{Holdemania} and had more abundant populations of
\emph{Lachnospiraceae} (Figure 5A). Next, we assessed the ability of the
community to predict moribundity. Bacteria grouped by order rank
classification was sufficient to predict which mice would succumb to the
infection before the end of the experiment (mean AUROC = 0.9205, Figure
S3). Many populations contributed to an increase odds of moribundity
(Figure 5B). Populations related to \emph{Bifidobacteriales} and
\emph{Clostridia} decreased the odds of a moribund outcome. Lastly, the
relative abundances of OTUs were able to predict a high or low
histopathologic score 10 dpi (histopathologic scores were dichotomized
as in previous analysis, mean AUROC = 0.99, Figure S3). The model
identified some similar OTUs as the LEfSe analysis, such as
\emph{Murimonas} (OTU 48), \emph{Bacteroides} (OTU 7), and
\emph{Hungatella} (OTU 24). These models have shown that the relative
abundance of bacterial populations and their relationship to each other
could be used to predict the variation in moribundity, histopathologic
score, and detectable toxin of CDI.

@@

Challenging mice colonized with different human fecal communities with
\emph{C. difficile} RT027 demonstrated that variation in members of the
gut microbiome affects \emph{C. difficile} infection disease severity.
Our analysis revealed an association between the relative abundance of
bacterial community members and disease severity. Previous studies
investigating the severity of CDI disease involving the microbiome have
had limited ability to interrogate this relationship between the
microbiome and disease severity. Studies that have used clinical data
have limited ability to control variation in the host, microbiome or
\emph{C. difficile} ribotype (23). Murine experiments typically use a
single mouse colony and different \emph{C. difficile} ribotypes to
create severity differences (24). Recently, our group has begun
uncovering the effect microbiome variation has on \emph{C. difficile}
infection. We showed the variation in the bacterial communities between
mice from different mouse colonies resulted in different clearance rates
of \emph{C. difficile} (16). We also showed varied ability of mice to
spontaneously eliminate \emph{C. difficile} infection when they were
treated with different antibiotics prior to \emph{C. difficile}
challenge (25). Overall, the results presented here have demonstrated
that the gut bacterial community contributed to the severity of \emph{C.
difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 26, 27). Those levels are driven by the
activities in the intestine (28). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (20, 29--32).
However, other studies have shown that disease is not necessarily linked
with toxin production (33) or the strain (34). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (35--38). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
24, 39, 40). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy murine gut
community and is only able to become established after a perturbation
(21). Once colonized, the different communities lead to different
metabolic responses and dynamics of the \emph{C. difficile} population
(9, 25, 41). Gut bacteria metabolize primary bile acids into secondary
bile acids (4, 42, 43). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (10, 11, 44,
45). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (46, 47). The nutrient environment
affects toxin production (48). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

A myriad of studies have explored the relationship between the
microbiome and CDI disease. Studies examining difference in disease
often use different \emph{C. difficile} strains or ribotypes in mice
with similar microbiota as a proxy for variation in disease, such as
strain 630 for non-severe and RT027 for severe (20, 29, 30, 49). Studies
have also demonstrated variation in infection through tapering
antibiotic dosage (21, 25, 50) or by reducing the amount of \emph{C.
difficile} cells or spores used for the challenge (20, 50). These
studies often either lack variation in the initial microbiome or have
variation in the \emph{C. difficile} infection itself, confounding any
association between variation in severity and the microbiome. Recent
studies have shown variation in the initial microbiome, via different
murine colonies or colonizing germ-free mice with human feces, that were
challenged with \emph{C. difficile} resulted in varied outcomes of the
infection (15, 16, 51).

Our data have demonstrated gut bacterial relative abundances associate
with variation in toxin production, histopathologic scoring of the cecal
tissue and mortality. This analysis revealed populations of
\emph{Akkermansia}, \emph{Anaerotignum}, \emph{Blautia},
\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
more abundant in the microbiome of non-moribund mice which had low
histopathologic scores and no detected toxin. The protective role of
these bacteria are supported by previous studies. \emph{Blautia},
\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
involved in dietary fiber fermentation and associated with healthy
communities (52--54). \emph{Anaerotignum}, which produce short chain
fatty acids, has been associated with healthy communities (55, 56).
\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
abundant in mice which had a low histopathologic scores but have
contradictory supporting evidence in the current literature. In our
data, a population of \emph{Akkermansia}, OTU 5, was most abundant in
the non-moribund mice with low histopathologic scores but moribund mice
had increased population of \emph{Akkermansia}, OTU 8. This difference
could indicate either a more protective mucus layer was present
inhibiting colonization (57, 58) or mucus consumption by
\emph{Akkermansia} could have been crossfeeding \emph{C. difficile} or
exposing a niche for \emph{C. difficile} (59--61). Similarly,
\emph{Enterocloster} was more abundant and associated with low
histopathologic scores. It has been associated with healthy populations
and has been used to mono-colonize germ-free mice to reduce the ability
of \emph{C. difficile} to colonize (62, 63). However,
\emph{Enterocloster} has also been involved in infections, such as
bacteremia (64, 65). These data have exemplified populations of bacteria
that have the potential to be either protective or harmful. Thus, the
disease outcome is not likely based on the abundance of individual
populations of bacteria, rather it is the result of the interactions of
the community.
